\chapter{マッチャーを定義しよう}\label{matcher}

\section{Egison内部のパターンマッチアルゴリズム}

This section explains and implements the pattern-matching algorithm of Egison, which is described in detail in Sect. 5 and Sect. 7 of the original paper of Egison~\cite{egi2018Aplas}.
After the explanation of the pattern-matching algorithm, I introduce a simplified implementation of this algorithm in Scheme.
This implementation is called by the proposed macros whose implementation is introduced in Sect.~\ref{method}.

\medskip

First, I start by explaining the pattern-matching algorithm of Egison briefly.
In Egison, pattern matching is implemented as reductions of \emph{matching states}.
A matching state consists of a stack of \emph{matching atoms} and an intermediate result of pattern matching.
A matching atom is a triple of a pattern, target, and matcher.
In each step of a pattern-matching process, the top matching atom of the stack of matching atoms is popped off.
From this matching atom, a list of lists of next matching atoms is calculated.
Each list of the next matching atoms is pushed on the stack of matching atoms of the matching state.
As a result, a single matching state is reduced to multiple stacks of matching states in a single reduction step.
Pattern matching is recursively executed for each matching state.
When a stack becomes empty, it means pattern matching for this matching state succeeded.
%\new{
Fig.~\ref{fig:reductionPath} shows the reduction path when executing the \lstinline{match-all} expression below.
The matching state that is not grayed-out in the $i$-th row is reduced to matching states in the $(i+1)$-th row.
For example, the first matching state in the second row is reduced to the matching state in the third row.
\texttt{MState} takes two arguments, a stack of matching atoms and an intermediate pattern-matching result.
The value pattern \lstinline{,m} is transformed to \lstinline{(val ,(lambda (m) m))}.
This transformation is explained in Sect.~\ref{method-val-pat}.
%}%new

\begin{lstlisting}[language=egison]
(match-all '(2 8 2) (Multiset Integer) [(cons m (cons ,m _)) m]) ; (2 2)
\end{lstlisting}

\begin{figure}[t]
\begin{center}
  \footnotesize
  \begin{tabular}{ll}
    \toprule
    1 &  \texttt{(MState \{[(cons m (cons (val ,(lambda (m) m)) \_)) (Multiset Integer) \{2 8 2\}]\} \{\})} \\ \hline
    2 &  \thead[l]{\texttt{(MState \{[m Integer 2] [(cons (val ,(lambda (m) m)) \_) (Multiset Integer) \{8 2\}]\} \{\})} \\
                   \textcolor{gray}{\texttt{(MState \{[m Integer 8] [(cons (val ,(lambda (m) m)) \_) (Multiset Integer) \{2 2\}]\} \{\})}} \\
                   \textcolor{gray}{\texttt{(MState \{[m Integer 2] [(cons (val ,(lambda (m) m)) \_) (Multiset Integer) \{2 8\}]\} \{\})}}} \\ \hline
    3 &  \texttt{(MState \{[m Something 2] [(cons (val ,(lambda (m) m)) \_) (Multiset Integer) \{8 2\}]\} \{\})} \\ \hline
    4 &  \texttt{(MState \{[(cons (val ,(lambda (m) m)) \_) (Multiset Integer) \{8 2\}]\} \{2\})} \\ \hline
    5 &  \thead[l]{\textcolor{gray}{\texttt{(MState \{[,m Integer 8] [\_ (Multiset Integer) \{2\}]\} \{2\})}} \\
                   \texttt{(MState \{[,m Integer 2] [\_ (Multiset Integer) \{8\}]\} \{2\})}} \\ \hline
    6 &  \texttt{(MState \{[\_ (Multiset Integer) \{8\}]\} \{2\})} \\ \hline
    7 &  \texttt{(MState \{[\_ Something \{8\}]\} \{2\})} \\ \hline
    8 &  \texttt{(MState \{\} \{2\})} \\
    \bottomrule
  \end{tabular}
\end{center}
  \caption{Reduction path of matching states}
  \label{fig:reductionPath}
\end{figure}

\section{\lstinline{unorderedPair}マッチャーの定義}

\section{\lstinline{multiset}マッチャーの定義}

Matcher definition is the most technical part in pattern-match-oriented programming.
This section explains the method for defining a matcher by showing a matcher definition for multisets.
The program below shows a matcher definition for multisets.
The multiset matcher is the most basic nontrivial matcher.
This section explores the detail of this definition.

\begin{lstlisting}[language=egison]
multiset a = matcher
               [] as () with
                 \tgt -> case tgt of
                           [] -> [()]
                           _ -> []
               $ : $ as (a, multiset a) with
                 \tgt -> matchAll tgt as list a with
                           $hs ++ $x : $ts -> (x, hs ++ ts)
               #$val as () with
                 \tgt -> match (val, tgt) as (list a, multiset a) with
                           ([], []) -> [()]
                           ($x : $xs, #x : #xs) -> [()]
                           (_, _) -> []
               $ as something with
                 \tgt -> [tgt]
\end{lstlisting}

A matcher is defined using the \lstinline{matcher} expression.
The \lstinline{matcher} expression is a built-in syntax of Egison.
\lstinline{matcher} takes a collection of \emph{matcher clauses}.
A matcher clause is a triple of a \emph{primitive-pattern pattern}, a \emph{next-matcher expression}, and a \emph{next-target expression}.

A matcher is a kind of function that takes a pattern and target, and returns lists of the next \emph{matching atoms}.
A matching atom is a triple of a pattern, target, and matcher.
A primitive-pattern pattern matches a pattern.
Patterns that match with \emph{pattern holes} (\lstinline|$| inside primitive-pattern patterns) are next patterns.
A next-matcher expression returns next matchers.
A next-target expression is a function that takes a target and returns a list of next targets.
A matcher generates a list of the next matching atoms by combining next patterns, next matchers, and a list of next targets.\footnote{In Egison, pattern matching is implemented as reductions of stacks of matching atoms.
Each list of the next matching atoms returned by a matcher is pushed to the stack of matching atoms.
As a result, a single stack of matching atom is reduced to multiple stacks of matching atoms in a single reduction step.
Pattern matching is recursively executed for each stack of matching atoms.
When a stack becomes empty, it means pattern matching for this stack succeeded.}
p
The \lstinline{multiset} matcher has four matcher clauses.
The first matcher clause handles a nil pattern, and it checks whether the target is an empty collection or not.
The second matcher clause handles a cons pattern.
The third matcher clause handles a value pattern.
This matcher clause defines the equality of multisets.
The fourth matcher clause handles the other patterns for multiset: a pattern variable and wildcard.

\medskip

First, we focus on the second matcher clause.
The primitive-pattern pattern of the second matcher clause is \lstinline|$ : $|, and the next matcher expression is \lstinline|(a, multiset a)|.
It means two arguments of the cons pattern are next patterns and they are pattern-matched using the \lstinline|a| and \lstinline|multiset a| matchers, respectively.
\lstinline|a| is an argument of \lstinline{multiset} and the matcher for inner elements of a multiset.
In the next target expression, a simple join-cons pattern is used to decompose a target collection into an element and the rest collection.
For example, when the target is a collection \lstinline{[1,2,3]}, this next-target expression returns \lstinline{[(1,[2,3]),(2,[1,3]),(3,[1,2])]}.
Each tuple of the next targets is pattern-matched using the next patterns and the next matchers recursively.
For example, \lstinline{1} and \lstinline{[2,3]} are pattern-matched using the \lstinline{a} and \lstinline{multiset a} matcher with the first and the second argument of the cons pattern, respectively.

\medskip

Next, we focus on the third matcher clause.
This matcher clause is as technical as the second matcher clause.
The primitive-pattern pattern of this matcher clause is \lstinline|#$val|.
It is called a \emph{value-pattern pattern}.
A value-pattern pattern matches a value pattern.
This matcher clause compares the content of a value pattern (\lstinline{val}) and a target (\lstinline{tgt}) as multisets.
The \lstinline{match} expression is used for this comparison.
Interestingly, \lstinline|tgt| is recursively pattern-matched as \lstinline{multiset a}.

The first and the third match clauses of this \lstinline{match} expression are simple.
The first match clause states that it returns \lstinline|[()]| when both \lstinline{val} and \lstinline{tgt} are empty.
This return value means pattern matching for the value pattern succeeded.
The third match clause states that it returns \lstinline|[]| if pattern matching for the patterns of both the first and the second match clause failed.
This return value means pattern matching for the value pattern failed.

The second match clause is the most technical part of this \lstinline{match} expression.
The value pattern \lstinline|#xs| is recursively pattern-matched using this matcher clause itself.
The collection \lstinline{xs} is one element shorter than \lstinline{tgt}.
Therefore, this recursion finally reaches the first or the third match clause if \lstinline{val} and \lstinline{tgt} are finite.

\medskip

Finally, let us also explain the fourth matcher clause.
This matcher clause creates the next matching atom by just changing the matcher from \lstinline{multiset a} to \lstinline{something}.

\section{\lstinline{soretedList}マッチャーの定義}

Modularization of pattern-matching algorithms by matchers not by patterns enables polymorphic patterns.
However, its merit extends beyond polymorphic patterns; matchers enable descriptions of more efficient pattern matching keeping patterns concise.
The reason is that pattern matching against patterns inside matcher definitions allows us to describe more detailed pattern-matching algorithms.
This section shows such an example, a matcher for sorted lists.

The program that used a doubly-nested join-cons pattern for enumerating pairs of prime numbers whose forms are $(p,p+6)$ gets slower when the number of the enumerating prime pairs gets larger.
The reason is that the program enumerates all the combinations of prime numbers.
For example, the program tries to match all the pairs such as $(3,5), (3,7), (3,11), (3,13), (3, 17), (3,19)$, and so on.
However, we should avoid enumerating the pairs after $(3,11)$, the first pair whose difference is more than $6$.
This is because it is obvious that the difference between all the pairs after $(3,11)$ are more than $6$.

\begin{lstlisting}[language=egison]
take 10 (matchAll primes as sortedList integer with
           _ ++ $p : (_ ++  #(p + 6) : _) -> (p, p + 6))
-- [(5,11),(7,13),(11,17),(13,19),(17,23),(23,29),(31,37),(37,43),(41,47),(47,53)]
\end{lstlisting}

We can avoid this unnecessary search by creating a new matcher that is specialized for sorted lists.
We can define such a matcher by adding a matcher clause with the primitive-pattern pattern \lstinline|$ ++ #$px : $| to the \lstinline{list} matcher as shown below.
This matcher clause improves the theoretical time complexity of the above pattern from $O(n^2)$ to $O(n)$.

\begin{lstlisting}[language=egison]
sortedList a =
  matcher
    $ ++ #$px : $ as (sortedList a, sortedList a)
      \tgt -> matchAll tgt as list a with
                loop $i (1, $n) ((and ?(\x -> x < px) $h_i) : ...) (#px : $ts) ->
                  (map (\i -> h_i) [1..n], ts)
    ...
\end{lstlisting}

Note that this method is only applicable to Egison that modularizes pattern-matching methods for each matcher, not for each pattern.
The reason is that we need to match patterns whose form is \lstinline|_ ++ #x : _| as mentioned above.
If pattern-matching methods are modularized for each pattern, we need to introduce a new pattern constructor \lstinline|joinCons ... ...| that is equivalent to \lstinline|_ ++ ... : ...| for this purpose.
