\chapter{より実践的なパターンマッチ指向プログラミング}\label{practice}

パターンマッチ指向プログラミングがどのような場面で役に立つか現在わかっている範囲で議論する．
前章までの例は，すべて単一の\lstinline{matchAll}や\lstinline{match}で記述した例ばかりを紹介してきたが，本章ではより大きなプログラムの記述の中でEgisonのパターンマッチがどのように役に立つのか調べる．

\section{SATソルバー}\label{practice-sat}

To see the effect of pattern-match-oriented programming for implementing practical algorithms, we implement a SAT solver.
A SAT solver determines whether a given propositional logic formula has an assignment for which the formula evaluates to true.
Input formulae for SAT solvers are often in \emph{conjunctive normal form}.
A formula in conjunctive normal form is a conjunction of clauses, which are disjunctions of \emph{literals}.
A literal is a formula whose form is $p$ or $\neg p$.
For example, $(p \lor q) \wedge (\neg p \lor r) \wedge (\neg p \lor \neg r)$ is a formula in conjunctive normal form that has a solution; $p = \texttt{False}$, $q = \texttt{True}$, and $r = \texttt{True}$.

\subsection{The Davis-Putnam Algorithm}

In our implementation, propositional logic formulae in conjunctive normal form are represented as a collection of collections of literals.
We can pattern-match them as a multiset of multisets of literals because both $\wedge$ and $\lor$ are commutative operators.
Furthermore, we represent a literal as an integer.
We represent positive and negative literals as positive and negative integers respectively: for example, $p$ and $\neg p$ are represented as $1$ and $-1$, respectively.
Therefore, the matcher for these formulae can be defined by simply composing matchers as \lstinline{multiset (multiset integer)}.

The program below shows the main part of our implementation of the Davis-Putnam algorithm\cite{harrison2009handbook}.
The \lstinline{sat} function takes a list of propositional variables and a logical formula, and returns \lstinline{True} if there is a solution, otherwise returns \lstinline{False}.

\begin{lstlisting}[language=egison]
sat vars cnf = match (vars, cnf) as (multiset integer, multiset (multiset integer)) with
  (_       , []) -> True
  (_       , [] : _) -> False
  (_       , ($x : []) : _) -> -- 1-literal rule
    sat (delete (abs x) vars) (assignTrue x cnf)
  ($v : $vs, !((#(negate v): _) : _)) -> -- pure literal rule (positive)
    sat vs (assignTrue v cnf)
  ($v : $vs, !((#v: _) : _)) -> -- pure literal rule (negative)
    sat vs (assignTrue (negate v) cnf)
p  ($v : $vs, _) -> -- otherwise
    sat vs (deleteClausesWith [v, negate v] cnf ++ resolveOn v cnf)
\end{lstlisting}

The first match clause states that the input formula has a solution when it is empty.
The second match clause states that that there is no solution when clauses include an empty clause.
The third match clause represents \emph{1-literal rule}.
When the input formula includes a clause with a single literal, we can assign that literal \lstinline{True} at once.
The fourth match clause states that that if there is a propositional variable that appears only positively, we can set the value of this literal \lstinline{True} and remove the propositional variable of \lstinline{x} from the variable list and the clauses that include this literal from the formula.
For example, $(p \lor q) \wedge (\neg p \lor r) \wedge (\neg p \lor \neg r)$ contains a propositional variable $q$ only positively, so we can assign $q$ \lstinline{True} and remove the first clause from the next recursion.
The fifth match clause states that the opposite of the fourth match clause.
It removes the clauses that include this literal if there is a propositional variable that appears only negatively ($p$ in the above sample is such propositional variable).
The final match clause applies the resolution principle.
This match clause lists all pairs of clauses $p \lor C$ and $\neg p \lor D$ (let $C$ and $D$ be a disjunction of literals), and obtains new clauses $C \lor D$.

The above definition of \lstinline{sat} describes all rules for simplifying an input formula by fully utilizing pattern matching for multisets.
In traditional functional languages, we need to call several library functions and define several helper functions to describe these conditional branches.
We can compare this implementation with the same algorithm implemented in OCaml in~\cite{harrison2009handbook}.

\subsection{Pattern Matching for Resolution}

We can elegantly define the \lstinline{resolveOn} function using a sequential not-pattern.

First, let us show a naive implementation of \lstinline{resolveOn}.
The \lstinline{resolveOn} function is defined with a single \lstinline{matchAll} expression as follows.

\begin{lstlisting}[language=egison]
resolveOn v cnf = matchAll cnf as multiset (multiset integer) with
  (#v : $xs) : (#(negate v) : $ys) : _ -> unique (filter (\c -> not (tautology c)) (xs ++ ys))
\end{lstlisting}

\noindent
The pattern for enumerating the pair of clauses $p \lor C$ and $\neg p \lor D$ is described simply utilizing pattern-matching for a multiset of multisets.
Note that the above \lstinline{resolveOn} removes tautological clauses by using the \lstinline{tautology} predicate in the body of the match clause.

We can remove this use of the \lstinline{tautology} predicate by using a sequential not-pattern discussed in Sect.~\ref{tuple-of-collections}.
The sequential pattern is effectively used to describe that the literal \lstinline{x} appeared in \lstinline{xs} does not appear negatively in \lstinline{ys}.

\begin{lstlisting}[language=egison]
resolveOn v cnf = matchAll cnf as multiset (multiset integer) with
  [(#v : (and @ $xs)) : (#(negate v) : (and @ $ys)) : _,
   !($x : _, #(negate x) : _)] -> unique (xs ++ ys)
\end{lstlisting}

\subsection{Separating Two Kinds of Loops}

The SAT solver presented in this section is an important sample in the sense that it is the only sample that contains a loop that we cannot remove by pattern-match-oriented programming.
This unremovable loop is the recursion of the \lstinline{sat} function.
This recursion is essential for narrowing the search tree.
This narrowing is impossible by simple backtracking.
On the other hand, all the other loops that can be implemented in backtracking algorithms are pushed into the patterns.
In the traditional style, we describe the 1-literal rule and pure literal rules by using recursive functions such as \lstinline{find}, \lstinline{partition}, \lstinline{subtract} and \lstinline{intersect}~\cite{harrison2009handbook}.
Thus, the separation of two kinds of loops increases the readability of practical algorithms.

\section{ユニフィケーション}\label{practice-unify}

\section{ゲーム木のパターンマッチ}\label{practice-tree}

